---
output:
  github_document:
    pandoc_args: --webtex=https://chart.googleapis.com/chart?cht=tx&chl=
bibliography: README.bib
nocite: |
  @Polyak92
---

# Multivariate State Space Models
[![Build Status on Travis](https://travis-ci.org/boennecd/mssm.svg?branch=master,osx)](https://travis-ci.org/boennecd/mssm)

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  error = FALSE, cache = "./README-cache/", fig.path = "./README-fig/", 
  echo = TRUE)
options(digits = 4, scipen = 7)
.fig_height_small <- 4
source("render_toc.R")
palette(c("#000000", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00", 
          "#CC79A7", "#F0E442"))
```

This package provides methods to estimate models of the form 

$$y_{it} \sim g(\eta_{it}),\qquad i\in I_t$$ 
$$\eta_{it} = \vec\gamma^\top\vec x_{it} +\vec\beta_t^\top\vec z_{it}$$
$$\vec\beta_t = F\vec\beta_{t-1}+\vec\epsilon_t, \qquad \vec\epsilon_t\sim N(\vec 0, Q)$$

where $g$ is simple distribution, we observe $t=1,\dots,T$ periods, and $I_t$, 
$y_{it}$, $\vec x_{it}$, and 
$\vec z_{it}$ are known. What is multivariate is 
$\vec y_t = \{y_{it}\}_{i\in I_t}$ (though, $\vec \beta_t$ can also be 
multivariate) and this package is written to scale well 
in the dimension of $| I_t |$. The package uses independent 
particle filters as suggested by @Lin05. This particular type of filter 
can be used in the method suggested by @Poyiadjis11. I will show an example
of how to use the package through the rest of the document and highlight some 
implementation details. 

The package is not on CRAN but it can be installed from Github e.g., by 
calling

```{r eval = FALSE}
devtools::install_github("boennecd/mssm")
```


## Table of Contents

```{r echo = FALSE}
render_toc("README.Rmd", toc_header_name = "Table of Contents", toc_depth = 3L)
```

## Poisson Example

We simulate data as follows.

```{r simulate, fig.height = .fig_height_small}
# simulate path of state variables 
set.seed(78727269)
n_periods <- 312L
(F. <- matrix(c(.5, .1, 0, .8), 2L))
(Q <- matrix(c(.5^2, .1, .1, .7^2), 2L))
(Q_0 <- matrix(c(0.333, 0.194, 0.194, 1.46), 2L))

betas <- cbind(crossprod(chol(Q_0),        rnorm(2L)                      ), 
               crossprod(chol(Q  ), matrix(rnorm((n_periods - 1L) * 2), 2)))
betas <- t(betas)
for(i in 2:nrow(betas))
  betas[i, ] <- betas[i, ] + F. %*% betas[i - 1L, ]
par(mar = c(5, 4, 1, 1))
matplot(betas, lty = 1, type = "l")

# simulate observations
cfix <- c(-1, .2, .5, -1) # gamma
n_obs <- 100L
dat <- lapply(1:n_obs, function(id){
  x <- runif(n_periods, -1, 1)
  X <- cbind(X1 = x, X2 = runif(1, -1, 1))
  z <- runif(n_periods, -1, 1)
  
  eta <- drop(cbind(1, X, z) %*% cfix + rowSums(cbind(1, z) * betas))
  y <- rpois(n_periods, lambda = exp(eta))
  
  # randomly drop some
  keep <- .2 > runif(n_periods)
  
  data.frame(y = y, X, Z = z, id = id, time_idx = 1:n_periods)[keep, ]
})
dat <- do.call(rbind, dat)

# show some properties 
nrow(dat)
head(dat)
table(dat$y)

# quick smooth of number of events vs. time
par(mar = c(5, 4, 1, 1))
plot(smooth.spline(dat$time_idx, dat$y), type = "l", xlab = "Time", 
     ylab = "Number of events")

# and split by those with `Z` above and below 0
with(dat, {
  z_large <- ifelse(Z > 0, "large", "small")
  smooths <- lapply(split(cbind(dat, z_large), z_large), function(x){
    plot(smooth.spline(x$time_idx, x$y), type = "l", xlab = "Time", 
     ylab = paste("Number of events -", unique(x$z_large)))
  })
})
```

In the above, we simulate `r n_periods` (`n_periods`) with `r n_obs` (`n_obs`)
individuals. Each individual has a fixed covariate, `X2`, and two time-varying
covariates, `X1` and `Z`. One of the time-varying covariates, `Z`, has a 
random slope. Further, the intercept is also random. 

### Log-Likelihood Approximations

We start by estimating a generalized linear model without random effects.

```{r fit_glm}
glm_fit <- glm(y ~ X1 + X2 + Z, poisson(), dat)
summary(glm_fit)
logLik(glm_fit)
```

Next, we make a log-likelihood approximation with the implemented particle at 
the true parameters with the `mssm` function.

```{r fit_mssm}
library(mssm)
ll_func <- mssm(
  fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat, 
  # make it explict that there is an intercept (not needed)
  random = ~ 1 + Z, ti = time_idx, control = mssm_control(
    n_threads = 5L, N_part = 500L, what = "log_density"))

system.time(
  mssm_obj <- ll_func$pf_filter(
    cfix = cfix, disp = numeric(), F. = F., Q = Q))

# returns the log-likelihood approximation
logLik(mssm_obj)
```

We get a much larger log-likelihood as expected. We can plot the predicted 
values of state variables from the filter distribution.

```{r plot_filter, fig.height = .fig_height_small, fig.width = 12L}
# get predicted mean and prediction interval 
filter_means <- plot(mssm_obj, do_plot = FALSE)

# plot with which also contains the true paths
for(i in 1:ncol(betas)){
  be <- betas[, i]
  me <- filter_means$means[i, ]
  lb <- filter_means$lbs[i, ]
  ub <- filter_means$ubs[i, ]
  
  #     dashed: true paths
  # continuous: predicted mean from filter distribution 
  #     dotted: prediction interval
  par(mar = c(5, 4, 1, 1))
  matplot(cbind(be, me, lb, ub), lty = c(2, 1, 3, 3), type = "l", 
          col = "black", ylab = rownames(filter_means$lbs)[i])
}
```

We can get the effective sample size at each point in time with the `get_ess`
function.

```{r show_ess, fig.height = .fig_height_small}
(ess <- get_ess(mssm_obj))
plot(ess)
```

We can compare this what we get by using a so-called bootstrap (like) filter
instead.

```{r comp_boot, fig.height = .fig_height_small}
local({
  ll_boot <- mssm(
    fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat, 
    random = ~ Z, ti = time_idx, control = mssm_control(
      n_threads = 5L, N_part = 500L, what = "log_density", 
      which_sampler = "bootstrap"))
  
  print(system.time(
    boot_fit <- ll_boot$pf_filter(
      cfix = coef(glm_fit), disp = numeric(), F. = F., Q = Q)))
  
  plot(get_ess(boot_fit))
})
```

The above is not much faster (and maybe slower in this run) as the bulk of 
the computation is not in the sampling step. We can also compare the 
log-likelihood approximation with what we get if we choose parameters close 
to the GLM estimates.

```{r comp_close_glm}
mssm_glm <- ll_func$pf_filter(
  cfix = coef(glm_fit), disp = numeric(), F. = diag(1e-8, 2), 
  Q = diag(1e-4^2, 2))
logLik(mssm_glm)
```

### Parameter Estimation

We will need to estimate the parameters for real applications. We could do 
this e.g., with a Monte Carlo expectation-maximization algorithm or by using 
a Monte Carlo approximation of the gradient. Currently, the latter is only 
available and the user will have to write a custom function to perform 
the estimation. 
I will provide an example below. The `sgd` function is not a part of the 
package. Instead the package provides a way to approximate the gradient and 
allows the user to perform subsequent maximization (e.g., with constraints or 
penalties). The definition of the 
`sgd` is given at the end of this file as it is somewhat long. 

```{r mc_grad_est, echo = FALSE}
# Stochastic gradient descent for mssm object. The function assumes that the 
# state vector is stationary. The default values are rather arbitrary.
# 
# Args:
#   object: an object of class mssmFunc. 
#   n_it: number of iterations. 
#   lrs: learning rates to use. Must have n_it elements. Like problem specific. 
#   avg_start: index to start averaging. See Polyak et al. (1992) for arguments
#              for averaging.
#   cfix: starting values for fixed coefficients. 
#   F.: starting value for transition matrix in conditional distribution of the 
#       current state given the previous state.
#   Q: starting value for covariance matrix in conditional distribution of the 
#      current state given the previous. 
#   verbose: TRUE if output should be printed during estmation. 
# 
# Returns:
#   List with estimates and the log-likelihood approximation at each iteration. 
sgd <- function(
  object, n_it = 150L, 
  lrs = 1e-2 * (1:n_it)^(-1/2), avg_start = max(1L, as.integer(n_it * 4L / 5L)),
  cfix, F., Q,  verbose = FALSE)
{
  # make checks
  stopifnot(
    inherits(object, "mssmFunc"), object$control$what == "gradient", n_it > 0L, 
    all(lrs > 0.), avg_start > 1L, length(lrs) == n_it)
  n_fix <- nrow(object$X)
  n_rng <- nrow(object$Z)
  
  # objects for estimates at each iteration and log-likelihood approximations
  ests <- matrix(
    NA_real_, n_it + 1L, n_fix + n_rng * n_rng + n_rng * (n_rng  + 1L) / 2L)
  ests[1L, ] <- c(cfix, F., Q[lower.tri(Q, diag = TRUE)])
  lls <- rep(NA_real_, n_it)
  
  # indices of the different components
  idx_fix <- 1:n_fix
  idx_F   <- 1:(n_rng * n_rng) + n_fix
  idx_Q   <- 1:(n_rng * (n_rng  + 1L) / 2L) + n_fix + n_rng * n_rng
  
  
  # function to set the parameters
  library(matrixcalc) # TODO: get rid of this
  dup_mat <- duplication.matrix(ncol(Q))
  set_parems <- function(i){
    # select whether or not to average
    idx <- if(i > avg_start) avg_start:i else i
    
    # set new parameters
    cfix <<-             colMeans(ests[idx, idx_fix, drop = FALSE])
    F.[] <<-             colMeans(ests[idx, idx_F  , drop = FALSE])
    Q[]  <<- dup_mat %*% colMeans(ests[idx, idx_Q  , drop = FALSE])
    
  }
    
  # run algorithm
  max_half <- 25L
  for(i in 1:n_it + 1L){
    # get gradient. First, run the particle filter
    filter_out <- object$pf_filter(
      cfix = cfix, disp = numeric(), F. = F., Q = Q, seed = NULL)
    lls[i - 1L] <- c(logLik(filter_out))
    
    # then get the gradient associated with each particle and the log 
    # normalized weight of the particles
    grads <- tail(filter_out$pf_output, 1L)[[1L]]$stats
    ws    <- tail(filter_out$pf_output, 1L)[[1L]]$ws_normalized
    
    # compute the gradient and take a small step
    grad <- colSums(t(grads) * drop(exp(ws)))
    lr_i <- lrs[i - 1L]
    k <- 0L
    while(k < max_half){
      ests[i, ] <- ests[i - 1L, ] + lr_i * grad 
      set_parems(i)
      
      # check that Q is positive definite and the system is stationary
      c1 <- all(abs(eigen(F.)$values) < 1)
      c2 <- all(eigen(Q)$values > 0)
      if(c1 && c2)
        break
      
      # decrease learning rate
      lr_i <- lr_i * .5
      k <- k + 1L
    }
    
    # check if we failed to find a value within our constraints
    if(k == max_half)
      stop("failed to find solution within constraints")
    
    # print information if requested 
    if(verbose){
      cat(sprintf(
        "\nIt %5d: log-likelihood (current, max) %12.2f, %12.2f\n", 
        i - 1L, logLik(filter_out), max(lls, na.rm = TRUE)), 
          rep("-", 66), "\n", sep = "")
      cat("cfix\n")
      print(cfix)
      cat("F\n")
      print(F.)
      cat("Q\n")
      print(Q)
      cat(sprintf("Gradient norm: %10.4f\n", norm(t(grad))))
      print(get_ess(filter_out))
      
    }
  } 
  
  list(estimates = ests, logLik = lls, F. = F., Q = Q, cfix = cfix)
}

# Stochastic gradient descent for mssm object using the Adam algorithm. The  
# function assumes that the state vector is stationary.
# 
# Args:
#   object: an object of class mssmFunc. 
#   n_it: number of iterations. 
#   mp: decay rate for first moment.
#   vp: decay rate for secod moment.
#   lr: learning rate.
#   cfix: starting values for fixed coefficients. 
#   F.: starting value for transition matrix in conditional distribution of the 
#       current state given the previous state.
#   Q: starting value for covariance matrix in conditional distribution of the 
#      current state given the previous. 
#   verbose: TRUE if output should be printed during estmation. 
# 
# Returns:
#   List with estimates and the log-likelihood approximation at each iteration. 
adam <- function(
  object, n_it = 150L, mp = .9, vp = .999, lr = .01, cfix, F., Q, 
  verbose = FALSE)
{
  # make checks
  stopifnot(
    inherits(object, "mssmFunc"), object$control$what == "gradient", n_it > 0L,
    lr > 0., mp > 0, mp < 1, vp > 0, vp < 1)
  n_fix <- nrow(object$X)
  n_rng <- nrow(object$Z)

  # objects for estimates at each iteration and log-likelihood approximations
  ests <- matrix(
    NA_real_, n_it + 1L, n_fix + n_rng * n_rng + n_rng * (n_rng  + 1L) / 2L)
  ests[1L, ] <- c(cfix, F., Q[lower.tri(Q, diag = TRUE)])
  lls <- rep(NA_real_, n_it)

  # indices of the different components
  idx_fix <- 1:n_fix
  idx_F   <- 1:(n_rng * n_rng) + n_fix
  idx_Q   <- 1:(n_rng * (n_rng  + 1L) / 2L) + n_fix + n_rng * n_rng

  # function to set the parameters
  library(matrixcalc) # TODO: get rid of this
  dup_mat <- duplication.matrix(ncol(Q))
  set_parems <- function(i){
    cfix <<-             ests[i, idx_fix]
    F.[] <<-             ests[i, idx_F  ]
    Q[]  <<- dup_mat %*% ests[i, idx_Q  ]

  }

  # run algorithm
  max_half <- 25L
  m <- NULL
  v <- NULL
  failed <- FALSE
  for(i in 1:n_it + 1L){
    # get gradient. First, run the particle filter
    filter_out <- object$pf_filter(
      cfix = cfix, disp = numeric(), F. = F., Q = Q, seed = NULL)
    lls[i - 1L] <- c(logLik(filter_out))

    # then get the gradient associated with each particle and the log
    # normalized weight of the particles
    grads <- tail(filter_out$pf_output, 1L)[[1L]]$stats
    ws    <- tail(filter_out$pf_output, 1L)[[1L]]$ws_normalized

    # compute the gradient and take a small step
    grad <- colSums(t(grads) * drop(exp(ws)))

    m <- if(is.null(m)) (1 - mp) * grad   else mp * m + (1 - mp) * grad
    v <- if(is.null(v)) (1 - vp) * grad^2 else vp * v + (1 - vp) * grad^2
    mh <- m / (1 - mp^(i - 1))
    vh <- v / (1 - vp^(i - 1))
    de <- mh / sqrt(vh + 1e-8)

    lr_i <- lr
    k <- 0L
    while(k < max_half){
      ests[i, ] <- ests[i - 1L, ] + lr_i * de
      set_parems(i)

      # check that Q is positive definite and the system is stationary
      c1 <- all(abs(eigen(F.)$values) < 1)
      c2 <- all(eigen(Q)$values > 0)
      if(c1 && c2)
        break

      # decrease learning rate
      lr_i <- lr_i * .5
      k <- k + 1L
    }

    # check if we failed to find a value within our constraints
    if(k == max_half){
      warning("failed to find solution within constraints")
      failed <- TRUE
      break
    }

    # print information if requested
    if(verbose){
      cat(sprintf(
        "\nIt %5d: log-likelihood (current, max) %12.2f, %12.2f\n",
        i - 1L, logLik(filter_out), max(lls, na.rm = TRUE)),
        rep("-", 66), "\n", sep = "")
      cat("cfix\n")
      print(cfix)
      cat("F\n")
      print(F.)
      cat("Q\n")
      print(Q)
      cat(sprintf("Gradient norm: %10.4f\n", norm(t(grad))))
      print(get_ess(filter_out))

    }
  }

  list(estimates = ests, logLik = lls, F. = F., Q = Q, cfix = cfix,
       failed = failed)
}
```

```{r use_sgd, cache = 1}
# setup mssmFunc object to use
ll_func <- mssm(  
  fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat, 
  random = ~ Z, ti = time_idx, control = mssm_control(
    n_threads = 5L, N_part = 200L, what = "gradient"))

# use stochastic gradient descent with averaging
set.seed(25164416)
system.time( 
  res <- sgd(
    ll_func, F. = diag(.5, 2), Q = diag(1, 2), cfix = coef(glm_fit), 
    lrs = .001 * (1:150L)^(-1/2)))

# use Adam algorithm instead
set.seed(25164416)
system.time( 
  resa <- adam(
    ll_func, F. = diag(.5, 2), Q = diag(2, 2), cfix = coef(glm_fit), 
    lr = .1))
```

A plot of the approximate log-likelihoods at each iteration is shown below 
along with the final estimates. 

```{r show_use_sgd, fig.height = .fig_height_small}
print(tail(res$logLik), digits = 6) # final log-likelihood approximations
par(mar = c(5, 4, 1, 1))
plot(     res$logLik       , type = "l")
plot(tail(res$logLik, 100L), type = "l") # only the final iterations

# final estimates
res$F. 
res$Q
res$cfix

# compare with output from Adam algorithm
print(tail(resa$logLik), digits = 6) # final log-likelihood approximations
plot(resa$logLik       , type = "l")
resa$F. 
resa$Q
resa$cfix
```

We may want to use more particles towards the end when we estimate the 
parameters. To do, we use the approximation described in the next section 
at the final estimates that we arrived at before. 

```{r cont_est, cache = 1, dependson = "use_sgd"}
ll_func <- mssm(
  fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat, 
  random = ~ Z, ti = time_idx, control = mssm_control(
    n_threads = 5L, N_part = 10000L, what = "gradient",
    which_ll_cp = "KD", aprx_eps = .01))

set.seed(25164416)
system.time( 
  res_final <- sgd(
    ll_func, F. = resa$F., Q = resa$Q, cfix = resa$cfix, 
    lrs = rep(.0001, 25L), n_it = 25L))
```

```{r show_cont_est}
plot(res_final$logLik, type = "l")
res_final$F. 
res_final$Q
res_final$cfix
```

```{r clean_param_ests, echo = FALSE}
# do this to not use them later by an error
rm(res, resa)
```

### Faster Approximation
One drawback with the particle filter we use is that it has $\mathcal{O}(N^2)$ 
computational complexity where $N$ is the number of particles. We can see 
this by changing the number of particles. 

```{r comp_w_n_part, cache = 1}
local({
  # assign function that returns a function that uses a given number of 
  # particles
  func <- function(N){
    ll_func <- mssm(
      fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat, 
      random = ~ Z, ti = time_idx, control = mssm_control(
        n_threads = 5L, N_part = N, what = "log_density"))
    function()
      ll_func$pf_filter(
        cfix = coef(glm_fit), disp = numeric(), F. = diag(1e-8, 2), 
        Q = diag(1e-4^2, 2))
      
  }
  
  f_100  <- func( 100L)
  f_200  <- func( 200L)
  f_400  <- func( 400L)
  f_800  <- func( 800L)
  f_1600 <- func(1600L)
  
  # benchmark. Should ĩncrease at ~ N^2 rate
  microbenchmark::microbenchmark(
    `100` = f_100(), `200` = f_200(), `400` = f_400(), `800` = f_800(),
    `1600` = f_1600(), times = 3L)
})
```

A solution is to use the dual k-d tree method I cover later. The computational
complexity is $\mathcal{O}(N \log N)$ for this method which is somewhat 
indicated by the run times shown below.

```{r KD_comp_w_n_part, cache = 1}
local({
  # assign function that returns a function that uses a given number of 
  # particles
  func <- function(N){
    ll_func <- mssm(
      fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat, 
      random = ~ Z, ti = time_idx, control = mssm_control(
        n_threads = 5L, N_part = N, what = "log_density", 
        which_ll_cp = "KD", KD_N_max = 6L, aprx_eps = 1e-2))
    function()
      ll_func$pf_filter(
        cfix = coef(glm_fit), disp = numeric(), F. = diag(1e-8, 2), 
        Q = diag(1e-4^2, 2))
      
  }
  
  f_100   <- func(  100L)
  f_200   <- func(  200L)
  f_400   <- func(  400L)
  f_800   <- func(  800L)
  f_1600  <- func( 1600L)
  f_12800 <- func(12800L) # <-- much larger
  
  # benchmark. Should increase at ~ N log N rate
  microbenchmark::microbenchmark(
    `100` = f_100(), `200` = f_200(), `400` = f_400(), `800` = f_800(), 
    `1600` = f_1600(), `12800` = f_12800(), times = 3L)
})
```

The `aprx_eps` controls the size of the error. To be precise about what this
value does then we need to some notation for the complete likelihood 
(the likelihood where we observe $\vec\beta_1,\dots,\vec\beta_T$s). This is

$$L = \mu_1(\vec \beta_1)g_1(\vec y_1 \mid \vec \beta_1)\prod_{t=2}^Tf(\vec\beta_t \mid\vec\beta_{t-1})g_t(y_t\mid\beta_t)$$

where $g_t$ is conditional distribution $\vec y_t$ given $\vec\beta_t$, $f$ is 
the conditional distribution of $\vec\beta_t$ given $\vec\beta_{t-1}$, and
$\mu$ is the time-invariant distribution of $\vec\beta_t$. 
Let $w_t^{(j)}$ be the weight of particle 
$j$ at time $t$ and $\vec \beta_t^{(j)}$ be the $j$th particle at time $t$. 
Then we ensure the error in our evaluation of terms 
$w_{t-1}^{(j)}f(\vec\beta_t^{(i)} \mid \vec\beta_{t-1}^{(j)})$ never 
exceeds

$$w_{t-1} \frac{u - l}{(u + l)/2}$$
where $u$ and $l$ are respectively an upper and lower bound of 
$f(\vec\beta_t^{(i)} \mid \vec\beta_{t-1}^{(j)})$. 
The question is how big the error is. 
Thus, we consider the error in the log-likelihood approximation at the 
true parameters. 

```{r comp_all_vs_aprx, cache = 1, fig.height = .fig_height_small, message = FALSE}
ll_compare <- local({
  N_use <- 500L 
  # we alter the seed in each run. First, the exact method
  ll_no_approx <- sapply(1:200, function(seed){
    ll_func <- mssm(
      fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat,
      random = ~ Z, ti = time_idx, control = mssm_control(
        n_threads = 5L, N_part = N_use, what = "log_density", 
        seed = seed))
    
    logLik(ll_func$pf_filter(
      cfix = cfix, disp = numeric(), F. = F., Q = Q))
  })
  
  # then the approximation
  ll_approx <- sapply(1:200, function(seed){
    ll_func <- mssm(
      fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat,
      random = ~ Z, ti = time_idx, control = mssm_control(
        n_threads = 5L, N_part = N_use, what = "log_density", 
        KD_N_max = 6L, aprx_eps = 1e-2, seed = seed, 
        which_ll_cp = "KD"))
    
    logLik(ll_func$pf_filter(
      cfix = cfix, disp = numeric(), F. = F., Q = Q))
  })
  
  list(ll_no_approx = ll_no_approx, ll_approx = ll_approx)
})
```

```{r show_comp_arell_aprx, fig.height = .fig_height_small}
par(mar = c(5, 4, 1, 1))
hist(
  ll_compare$ll_no_approx, main = "", breaks = 20L, 
  xlab = "Log-likelihood approximation -- no aprox")
hist(
  ll_compare$ll_approx   , main = "", breaks = 20L, 
  xlab = "Log-likelihood approximation -- aprox")
```

We can make a t-test for whether there is a difference between the two 
log-likelihood estimates

```{r t_test_comp_ll}
with(ll_compare, t.test(ll_no_approx, ll_approx))
```

The fact that there may only be a small difference if any is nice because 
now we can get a much better approximation (in terms of variance) quickly 
of e.g., the log-likelihood as shown below.

```{r show_ll_quick, cache = 1, message = FALSE}
ll_approx <- sapply(1:10, function(seed){
  N_use <- 10000L # many more particles
  
  ll_func <- mssm(
    fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat,
    random = ~ Z, ti = time_idx, control = mssm_control(
      n_threads = 5L, N_part = N_use, what = "log_density", 
      KD_N_max = 100L, aprx_eps = 1e-2, seed = seed, 
      which_ll_cp = "KD"))
  
  logLik(ll_func$pf_filter(
    cfix = cfix, disp = numeric(), F. = F., Q = Q))
}) 

# approximate log-likelihood
sd(ll_approx)
print(mean(ll_approx), digits = 6)

# compare sd with 
sd(ll_compare$ll_no_approx)
print(mean(ll_compare$ll_no_approx), digits = 6)
```

### Approximate Observed Information Matrix

Next, we look at approximating the observed information matrix with the method 
suggested by @Poyiadjis11.

```{r apprx_obs_info, cache = 1, dependson=c("use_sgd", "cont_est")}
ll_func <- mssm(
  fixed = y ~ X1 + X2 + Z, family = poisson(), data = dat,
  random = ~ Z, ti = time_idx, control = mssm_control(
    n_threads = 5L, N_part = 10000L, what = "Hessian",
    which_ll_cp = "KD", aprx_eps = .01))

set.seed(46658529)
system.time(
  mssm_obs_info <- ll_func$pf_filter(
    cfix = res_final$cfix, disp = numeric(), F. = res_final$F., 
    Q = res_final$Q))
```

We define a function below to get the gradient and approximate the observed 
information matrix from the returned output. Then we compare the output to 
the GLM we estimated and to the true parameters.

```{r show_apprx_obs_info}
# Function to subtract the gradient elements and diagonal block  
# matrices of the approximate observed information matrix. 
# 
# Args:
#   object: an object of class mssm.
#
# Returns:
#   approximate gradient elements and diagonal block matrices of the 
#   approximate observed information matrix. 
get_grad_n_obs_info <- function(object){
  stopifnot(inherits(object, "mssm"))

  # get dimension of the different components
  n_fix <- nrow(object$X)
  n_rng <- nrow(object$Z)
  dim_fix <- n_fix
  dim_F   <- n_rng * n_rng
  dim_Q   <- (n_rng * (n_rng  + 1L)) / 2L


  # get quantities for each particle
  quants <- tail(object$pf_output, 1L)[[1L]]$stats
  ws     <- tail(object$pf_output, 1L)[[1L]]$ws_normalized

  # get mean estimates
  meas <- colSums(t(quants) * drop(exp(ws)))

  # separate out the different components. Start with parameters for
  # the conditional density of the outcome given the state vector
  idx <- 1:n_fix
  dg  <- meas[idx]
  idx <- max(idx) + 1:(n_fix * n_fix)
  ddg <- matrix(meas[idx], n_fix, n_fix)

  # then the parameters for the conditional density of the state vector given
  # the previous
  idx <- max(idx) + 1:(dim_F + dim_Q)
  df <- meas[idx]
  idx <- max(idx) + 1:((dim_F + dim_Q) * (dim_F + dim_Q))
  ddf <- matrix(meas[idx], dim_F + dim_Q, dim_F + dim_Q)

  list(dg = dg, ddg = ddg, df = df, ddf = ddf)
}

# use function
out <- get_grad_n_obs_info(mssm_obs_info)

# Look at output for parameters in the observational equation
out$dg # approximate gradient
(cvov_fit <- solve(-out$ddg)) # inverse negative observed information matrix
(ses <- sqrt(diag(cvov_fit))) # approximate standard errors

# compare with glm model
(cvov_fit <- vcov(glm_fit))
sqrt(diag(cvov_fit))

# and relative to true parameters vs. estimated
rbind(true = cfix, glm = coef(glm_fit), mssm = res_final$cfix, se = ses)

# next look at parameters in state equation
out$df # approximate gradient
(cvov_fit <- solve(-out$ddf)) # inverse negative observed information matrix
(ses <- sqrt(diag(cvov_fit))) # approximate standard errors

# first four are for F. Estimate vs. true values are given below
rbind(true = c(F.), mssm = c(res_final$F.), se = ses[1:4])

# next three are w.r.t. the lower diagonal part of Q
rbind(
  true =           Q[lower.tri(Q, diag = TRUE)], 
  mssm = res_final$Q[lower.tri(Q, diag = TRUE)], 
  se = ses[5:7])
```

## Supported Families
The following families are supported:

 - The binomial distribution is supported with logit, probit, and cloglog link. 
 - The Poisson distribution is supported with square root and log link. 
 - The gamma distribution is supported with log link.

## Fast Sum-Kernel Approximation

This package contains a simple implementation of the dual-tree method like the one 
suggested by @Gray03 and shown in @Klaas06. The problem we want to solve is
the sum-kernel problem in @Klaas06. Particularly, we consider the situation 
where we have $1,\dots,N_q$ query particles denoted by 
$\{\vec Y_i\}_{i=1,\dots,N_q}$ and $1,\dots,N_s$ source particles denoted by
$\{\vec X_j\}_{j=1,\dots,N_s}$. For each query particle, we want to compute 
the weights 

$$W_i = \frac{\tilde W_i}{\sum_{k = 1}^{N_q} \tilde W_i},\qquad \tilde W_i = \sum_{j=1}^{N_s} \bar W_j K(\vec Y_i, \vec X_j)$$

where each source particle has an associated weight $\bar W_j$ and $K$ is a
kernel. Computing the above is $\mathcal{O}(N_sN_q)$ which is major 
bottleneck if $N_s$ and $N_q$ is large. However, one can use a 
[k-d tree](https://en.wikipedia.org/wiki/K-d_tree) for the query particles
and source particles and exploit that:

- $W_j K(\vec Y_i, \vec X_j)$ is almost zero for some pairs of nodes in the 
two k-d trees. 
- $K(\cdot, \vec X_j)$ is almost identical for some nodes in the k-d tree 
for the source particles. 

Thus, a substantial amount of computation can be skipped or approximated 
with e.g., the centroid in the source node with only a minor
loss of precision. The dual-tree approximation method is 
$\mathcal{O}(N_s\log N_s)$ and $\mathcal{O}(N_q\log N_q)$.
We start by defining a function to simulate the source 
and query particles (we will let the two sets be identical for simplicity). 
Further, we plot one draw of simulated points and illustrate the leafs in 
the k-d tree.

```{r sim_func}
######
# define function to simulate data
mus <- matrix(c(-1, -1, 
                 1, 1, 
                -1, 1), 3L, byrow = FALSE)
mus <- mus * .75
Sig <- diag(c(.5^2, .25^2))

get_sims <- function(n_per_grp){
  # simulate X
  sims <- lapply(1:nrow(mus), function(i){
    mu <- mus[i, ]
    X <- matrix(rnorm(n_per_grp * 2L), nrow = 2L)
    X <- t(crossprod(chol(Sig), X) + mu)
    
    data.frame(X, grp = i)
  })
  sims <- do.call(rbind, sims)
  X <- t(as.matrix(sims[, c("X1", "X2")]))
  
  # simulate weights
  ws <- exp(rnorm(ncol(X)))
  ws <- ws / sum(ws)
  
  list(sims = sims, X = X, ws = ws)
}

#####
# show example 
set.seed(42452654)
invisible(list2env(get_sims(5000L), environment()))

# plot points
par(mar = c(5, 4, 1, 1))
plot(as.matrix(sims[, c("X1", "X2")]), col = sims$grp + 1L)

# find k-d tree and add borders 
out <- mssm:::test_KD_note(X, 50L)
out$indices <- out$indices + 1L
n_ele <- drop(out$n_elems)
idx <- mapply(`:`, cumsum(c(1L, head(n_ele, -1))), cumsum(n_ele))
stopifnot(all(sapply(idx, length) == n_ele))
idx <- lapply(idx, function(i) out$indices[i])
stopifnot(!anyDuplicated(unlist(idx)), length(unlist(idx)) == ncol(X))

grps <- lapply(idx, function(i) X[, i])

borders <- lapply(grps, function(x) apply(x, 1, range))
invisible(lapply(borders, function(b) 
  rect(b[1, "X1"], b[1, "X2"], b[2, "X1"], b[2, "X2"])))
```

Next, we compute the run-times for the previous examples and compare the 
approximations of the un-normalized log weights, $\log \tilde W_i$, and 
normalized weights, $W_i$. The `n_threads` sets the number of threads to 
use in the methods.

```{r comp_run_times, cache = 1}
# run-times
microbenchmark::microbenchmark(
  `dual tree 1` = mssm:::FSKA (X = X, ws = ws, Y = X, N_min = 10L, 
                               eps = 5e-3, n_threads = 1L),
  `dual tree 4` = mssm:::FSKA (X = X, ws = ws, Y = X, N_min = 10L, 
                               eps = 5e-3, n_threads = 4L),
  `naive 1`     = mssm:::naive(X = X, ws = ws, Y = X, n_threads = 1L),
  `naive 4`     = mssm:::naive(X = X, ws = ws, Y = X, n_threads = 4L),
  times = 10L)

# The functions return the un-normalized log weights. We first compare
# the result on this scale
o1 <- mssm:::FSKA  (X = X, ws = ws, Y = X, N_min = 10L, eps = 5e-3, 
                    n_threads = 1L)
o2 <- mssm:::naive(X = X, ws = ws, Y = X, n_threads = 4L)

all.equal(o1, o2)
par(mar = c(5, 4, 1, 1))
hist((o1 - o2)/ abs((o1 + o2) / 2), breaks = 50, main = "", 
     xlab = "Delta un-normalized log weights")

# then we compare the normalized weights
func <- function(x){
  x_max <- max(x)
  x <- exp(x - x_max)
  x / sum(x)
}

o1 <- func(o1)
o2 <- func(o2)
all.equal(o1, o2)
hist((o1 - o2)/ abs((o1 + o2) / 2), breaks = 50, main = "", 
     xlab = "Delta normalized log weights")
```

Finally, we compare the run-times as function of $N = N_s = N_q$. The dashed 
line is "naive" method, the continuous line is the dual-tree method, and the 
dotted line is dual-tree method using 1 thread.

```{r run_times_N, cache = 1}
Ns <- 2^(7:14)
run_times <- lapply(Ns, function(N){ 
  invisible(list2env(get_sims(N), environment()))
  microbenchmark::microbenchmark(
    `dual-tree`   = mssm:::FSKA (X = X, ws = ws, Y = X, N_min = 10L, eps = 5e-3, 
                                 n_threads = 4L),
    naive         = mssm:::naive(X = X, ws = ws, Y = X, n_threads = 4L),
    `dual-tree 1` = mssm:::FSKA (X = X, ws = ws, Y = X, N_min = 10L, eps = 5e-3, 
                                 n_threads = 1L), 
    times = 5L)
}) 

Ns_xtra <- 2^(15:19) 
run_times_xtra <- lapply(Ns_xtra, function(N){
  invisible(list2env(get_sims(N), environment()))
  microbenchmark::microbenchmark(
    `dual-tree` = mssm:::FSKA (X = X, ws = ws, Y = X, N_min = 10L, eps = 5e-3, 
                               n_threads = 4L),
    times = 5L)
}) 
```
 
```{r plot_run_times_N}
library(microbenchmark)
meds <- t(sapply(run_times, function(x) summary(x, unit = "s")[, "median"]))
meds_xtra <- 
  sapply(run_times_xtra, function(x) summary(x, unit = "s")[, "median"])
meds <- rbind(meds, cbind(meds_xtra, NA_real_, NA_real_))
dimnames(meds) <- list(
  N = c(Ns, Ns_xtra) * 3L, method = c("Dual-tree", "Naive", "Dual-tree 1"))
meds
par(mar = c(5, 4, 1, 1))
matplot(c(Ns, Ns_xtra) * 3L, meds, lty = 1:3, type = "l", log = "xy", 
        ylab = "seconds", xlab = "N", col = "black")
```

## Function Definitions

```{r mc_grad_est, eval = FALSE}
```

# References
